---
title: "miRs 5 clusters, multivar"
author: "Niamh Errington"
date: "`r format(Sys.time(), '%d %B %Y')`" 
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: false
---

Create multivariable glm model for zscored data (keeping < 5 sd from mean). Clusters trained using training + interim. Models for each cluster trained on training set alone, using interim as test.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(kableExtra)
```

```{r, message=FALSE}
#load libraries
library(glmnet)
library(caret)
library(tidyverse)
library(dunn.test)
library(survival)
library(survminer)
library(readxl)
library(viridis)
library(pROC)
library(tidytext) #reorder_within
```

```{r, message=FALSE, warning=FALSE, cache=TRUE}
# Load previously cached chunks
#qwraps2::lazyload_cache_dir(path = "Z5_k5_PH12345_multivarmodels_cache/html")

#Read in data & create data set for each cluster
cluster_memberships <- read_csv( "PH12345_25memberships_Z5_allmiR.csv")

#load Zscored data
zscored <- read_csv("../../Data/z_TI_noout_5_PH12345.csv") 
zscored <- left_join(zscored, cluster_memberships, by = "SampleID")  

#load validation data  (doesn't have cluster assignments)
Validation <- read_csv("../../Data/z_V_noout_5_PH12345.csv") 

# load miRs in cipher
miRNAs_detected_in_Cipher_Serum <- read_excel("../../Data/miRNAs detected in Cipher Serum.xlsx")

miR_detect <- c("SampleID", "k5.memberships", "PARTITION", miRNAs_detected_in_Cipher_Serum$Serum.miR)

zscored <- zscored[, (colnames(zscored) %in% miR_detect)]


zscored <- zscored %>% filter(!is.na(k5.memberships) ) 

#Recode k5 memberships to letters
zscored$k5.memberships <- dplyr::recode(zscored$k5.memberships, `1` = "A", `2` = "B", `3` = "C", `4` = "D", `5`="E") %>% as.factor(.) %>% droplevels(.)

#Partition into training & interim based on original partitions
TrainingmiRs<- zscored %>% select(SampleID, PARTITION, k5.memberships, starts_with("hsa")) 
Training <- TrainingmiRs %>% filter(PARTITION == "TRAINING") %>% select(-PARTITION)
Interim <- TrainingmiRs %>% filter(PARTITION == "INTERIM") %>% select(-PARTITION)

```

```{r}
# Split the data into features (X) and target (y)
X <- Training %>% select(-SampleID, -k5.memberships)
y <- Training$k5.memberships

# Define train control for 10-fold cross-validation
train_control <- trainControl(
  method = "cv",
  number = 10, # 10-fold CV
  classProbs = TRUE, # Enable class probabilities 
  summaryFunction = multiClassSummary # Use a multi-class evaluation metric
)

# Train a multinomial logistic regression model using glmnet 
set.seed(123) # For reproducibility
multi_model <- train(
  x = X,
  y = y,
  method = "glmnet", # GLM with elastic net regularization
  family = "multinomial", # Multi-class classification
  trControl = train_control,
  tuneGrid = expand.grid(
    alpha = 1, #LASSO to reduce number of coefficients
    lambda = exp(seq(log(0.1), log(10), length = 100)) # higher lambda = stronger penalty & fewer coefficients
  )
)

# Check performance
#confusionMatrix(predict(multi_model, X), y)
#confusionMatrix(predict(multi_model, Interim), Interim$k5.memberships)


# Extract non-zero features
final_model <- multi_model$finalModel
best_lambda <- multi_model$bestTune$lambda

# Get coefficients for the best lambda
coef_matrix <- coef(final_model, s = best_lambda)

# Identify non-zero features
non_zero_features <- lapply(coef_matrix, function(class_matrix) {
  df <- as.data.frame(as.matrix(class_matrix))
  colnames(df) <- "Coefficient" 
  # Filter rows where the coefficient is non-zero 
  non_zero <- df %>%
    filter(Coefficient != 0) %>%
    rownames()
  
  return(non_zero)
})

X_reduced <- X[, colnames(X) %in% c(non_zero_features$A, non_zero_features$B, non_zero_features$C, non_zero_features$D, non_zero_features$E)]

int_reduced <- Interim[, colnames(Interim) %in% c(non_zero_features$A, non_zero_features$B, non_zero_features$C, non_zero_features$D, non_zero_features$E)]

# Retune model using only selected variables
set.seed(123) # For reproducibility
multi_model_reduced <- train(
  x = X_reduced,
  y = y,
  method = "glmnet", # GLM with elastic net regularization
  family = "multinomial", # Multi-class classification
  trControl = train_control,
  tuneGrid = expand.grid(
    alpha = 1, #LASSO to reduce number of coefficients
    lambda = exp(seq(log(0.01), log(10), length = 100)) # higher lambda = stronger penalty & fewer coefficients
  )
)

# Extract non-zero features
final_model <- multi_model_reduced$finalModel
best_lambda <- multi_model_reduced$bestTune$lambda

# Get coefficients for the best lambda
coef_matrix <- coef(final_model, s = best_lambda)

# Identify non-zero features
non_zero_features <- lapply(coef_matrix, function(class_matrix) {
  df <- as.data.frame(as.matrix(class_matrix))
  colnames(df) <- "Coefficient" 
  # Filter rows where the coefficient is non-zero 
  non_zero <- df %>%
    filter(Coefficient != 0) %>%
    rownames()
  
  return(non_zero)
})

#saveRDS(multi_model_reduced, "Omada/FinalModels/Multivar_model.rds")
#multi_model_reduced <- readRDS("Omada/FinalModels/Multivar_model.rds")
```

## Training set performance

```{r}
# Check performance
confusionMatrix(predict(multi_model_reduced, X_reduced), y)

# Get predicted probabilities on training data
predicted_probs_t <- predict(multi_model_reduced, X_reduced, type = "prob") # Probabilities for each class

# Compute multiclass AUC
auc_result_t <- multiclass.roc(y, predicted_probs_t)
auc_result_t

# Per-class AUCs
per_class_auc_t <- auc(auc_result_t)
per_class_auc_t
```

## Interim set performance

```{r}
confusionMatrix(predict(multi_model_reduced, int_reduced), Interim$k5.memberships)

# Get predicted probabilities on test data
predicted_probs_i <- predict(multi_model_reduced, int_reduced, type = "prob") # Probabilities for each class

# Compute multiclass AUC
auc_result_i <- multiclass.roc(Interim$k5.memberships, predicted_probs_i)
auc_result_i

# Per-class AUCs
per_class_auc_i <- auc(auc_result_i)
per_class_auc_i


# Compute ROC and AUC for each class
roc_list <- lapply(levels(Interim$k5.memberships), function(class) {
  roc(Interim$k5.memberships == class, predicted_probs_i[, class], quiet = TRUE)  # One-vs-all
})

# Extract AUC for each class
auc_values <- sapply(roc_list, function(roc_obj) auc(roc_obj))

# Prepare data for plotting
roc_data <- do.call(rbind, lapply(1:length(roc_list), function(i) {
  data.frame(
    TPR = roc_list[[i]]$sensitivities,   # True Positive Rate
    FPR = 1 - roc_list[[i]]$specificities, # False Positive Rate
    Class = levels(Interim$k5.memberships)[i]
  )
}))

# Plot the ROC curves for all classes
ggplot(roc_data, aes(x = FPR, y = TPR, color = Class)) +
  geom_line(size = 1.2) +
  theme_minimal() +
  labs(
    title = "ROC Curves for Multinomial Logistic Regression (Interim)",
    x = "False Positive Rate",
    y = "True Positive Rate",
    color = "Class"
  ) +
  annotate("text", x = 0.2, y = 0.2, 
           label = paste("AUCs:", paste(levels(Interim$k5.memberships), round(auc_values, 3), sep = " = ", collapse = ", ")),
           hjust = 0, size = 4)
```


```{r}
# Plot model performance 
ggplot(multi_model_reduced)

# Extract coefficients for the best lambda
coef_matrix <- coef(final_model, s = best_lambda)

# Combine coefficients into a data frame
coef_df <- do.call(rbind, lapply(seq_along(coef_matrix), function(class) {
  data <- as.data.frame(as.matrix(coef_matrix[[class]]))
  colnames(data) <- "Coefficient"
  data$Feature <- rownames(data)
  data$Class <- names(coef_matrix)[class]
  return(data)
}))

# Filter non-zero coefficients
coef_df <- coef_df[coef_df$Coefficient != 0, ]
```

## Coefficients

Length of coefficients for each cluster.

NB: You get 5 sets of coefficients in a multinomial glmnet model for 5 classes because the model constructs a unique linear predictor for each class. These predictors collectively determine the probabilities of each class for a given input using the softmax function.

```{r}
paste("cluster A miRNA:" ,length(unique(coef_df$Feature[coef_df$Class == "A"])) )
paste("cluster B miRNA:" ,length(unique(coef_df$Feature[coef_df$Class == "B"])) )
paste("cluster C miRNA:" ,length(unique(coef_df$Feature[coef_df$Class == "B"])) )
paste("cluster D miRNA:" ,length(unique(coef_df$Feature[coef_df$Class == "B"])) )
paste("cluster E miRNA:" ,length(unique(coef_df$Feature[coef_df$Class == "B"])) )

```

```{r}
#Fix miR names for plotting
coef_df$Feature <- gsub("\\.", "-", coef_df$Feature )
coef_df$Feature  <- gsub("\\(", "", coef_df$Feature )
coef_df$Feature  <- gsub("\\)", "", coef_df$Feature )

coef_df %>%
  mutate(Feature = reorder_within(Feature, abs(Coefficient), Class)) %>%
  ggplot(aes(x = Feature, y = Coefficient, fill = Class)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  coord_flip() +
  labs(title = "Non-Zero Coefficients by Feature and Class",
       x = "Feature",
       y = "Coefficient",
       fill = "Class") +
  facet_wrap(~Class, scales = "free", ncol = 1) +
  scale_x_reordered() + scale_fill_manual(values = c("#88CCEE", "#CC6677", "#DDCC77", "#332288", "#AA4499"))

coef_df %>%
  filter(Class %in% c("A", "B", "C")) %>%
  mutate(Feature = reorder_within(Feature, abs(Coefficient), Class)) %>%
  ggplot(aes(x = Feature, y = Coefficient, fill = Class)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  coord_flip() +
  labs(x = "Feature",
       y = "Coefficient",
       fill = "Class") +
  facet_wrap(~Class, scales = "free_y", ncol = 1) +
  scale_x_reordered() + scale_fill_manual(values = c("#88CCEE", "#CC6677", "#DDCC77"))

coef_df %>%
  filter(Class %in% c("C", "D", "E")) %>%
  mutate(Feature = reorder_within(Feature, abs(Coefficient), Class)) %>%
  ggplot(aes(x = Feature, y = Coefficient, fill = Class)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  coord_flip() +
  labs( x = "Feature",
       y = "Coefficient",
       fill = "Class") +
  facet_wrap(~Class, scales = "free_y", ncol = 1) +
  scale_x_reordered() + scale_fill_manual(values = c("#DDCC77", "#332288", "#AA4499"))


```

## AUC

```{r}


```

## Interim -A

```{r}
predictions_i_noA <- Interim %>% filter(k5.memberships != "A" )%>% select(-SampleID, -k5.memberships) %>% predict(multi_model, .)
confusionMatrix(predictions_i_noA, Interim$k5.memberships[Interim$k5.memberships != "A"] )

```

## Interim -B

```{r}
predictions_i_noB <- Interim %>% filter(k5.memberships != "B" )%>% select(-SampleID, -k5.memberships) %>% predict(multi_model, .)
confusionMatrix(predictions_i_noB, Interim$k5.memberships[Interim$k5.memberships != "B"] )

```

## Interim -C

```{r}
predictions_i_noC <- Interim %>% filter(k5.memberships != "C" )%>% select(-SampleID, -k5.memberships) %>% predict(multi_model, .)
confusionMatrix(predictions_i_noC, Interim$k5.memberships[Interim$k5.memberships != "C"] )

```

## Interim -D

```{r}
predictions_i_noD <- Interim %>% filter(k5.memberships != "D" )%>% select(-SampleID, -k5.memberships) %>% predict(multi_model, .)
confusionMatrix(predictions_i_noD, Interim$k5.memberships[Interim$k5.memberships != "D"] )

```

## Interim -E

```{r}
predictions_i_noE <- Interim %>% filter(k5.memberships != "E" )%>% select(-SampleID, -k5.memberships) %>% predict(multi_model, .)
confusionMatrix(predictions_i_noE, Interim$k5.memberships[Interim$k5.memberships != "E"] )

```

## Validation

```{r}
V <- Validation %>% select(starts_with("hsa")) 
# Re-align columns to match training data
V <- V[, colnames(X), drop = FALSE]
predictions_v <- predict(multi_model, V)

#add predictions to pheno data for Validation
Validation$clusterprediction <- predictions_v

#write csv
write_csv(Validation, "Validation_with_k5prediction.csv")

```

```{r}
as.data.frame(table(Validation$dana1, Validation$clusterprediction))  %>% group_by(Var2, Var1) %>% ggplot(aes(x=Var2, y=Freq, fill=Var1)) + geom_bar(stat="identity")  + scale_fill_manual(values = c("darkseagreen1","seagreen4", "grey", "lightblue", "midnightblue")) + ggtitle("WSPH PH Group, 5 clusters") + labs(x = "Cluster", y = "Number of patients", size = 14) +  guides(fill=guide_legend(title=NULL)) + theme(panel.background = element_blank(), panel.grid.major = element_line(colour = "grey"), axis.line = element_line(colour = "grey"), strip.text = element_text(size = 16)) +  guides(fill=guide_legend(title=NULL))

```


